---
output: github_document
author: "Erik Forsberg"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r setup, message = FALSE, include = FALSE}
library(D3mirt)
library(mirt)
knitr::knit_hooks$set(webgl = hook_webgl)
```

# `D3MIRT` Modeling

The `D3mirt` analysis is based on descriptive multidimensional item response theory (DMIRT; Reckase2009, 1985; Reckase & McKinley, 1991) and can be used to analyze dichotomous and polytomous items in a multidimensional ability space. In DMIRT analysis, also called within multidimensional modeling, it is assumed that items in a multidimensional ability space can measure single or multiple latent traits (Reckase, 2009, 1985; Reckase & McKinley, 1991). The method is said to be descriptive because the estimates describe item characteristics when more than one latent dimension is used in the analysis. The strength of the method is the possibility to represent item response functions visually with vector arrows located in a three-dimensional space.

The methodology is based on the two-parameter graded response model (GRM; Samejima, 1969) that is used to estimate multidimensional equivalents to the discrimination and difficulty parameters. The former becomes the multidimensional discrimination ($MDISC$) parameter and the latter the multidimensional difficulty ($MDIFF$) parameter.

The $MDIFF$ is interpreted similarly as the difficulty parameter in the unidimensional model, i.e., it describes a multidimensional distance from the origin to the item response function and shows the level of ability that is required for a higher response. Likewise, the $MDISC$ represents the highest level of discrimination an item can achieve when located in a multidimensional latent trait space. In addition, `D3mirt`, the $MDISC$ score is visualized by scaling the length of the vector arrows representing the item response functions.

# Overview

The package includes the following functions.

-   `modid()`: D3mirt Model Identification.
-   `D3mirt()`: 3D DMIRT Modeling.
-   `summary()`: summary for `D3mirt().`
-   `plotD3mirt`: Graphical Output for D3mirt.

## Installation

You can install the development version of D3mirt from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ForsbergPyschometrics/D3mirt")
```

In what follows, the `D3mirt` procedure will be described very briefly using the built-in data set "anes08_09offwaves". The data set ($N = 1046, M_{age} = 51.33, SD = 14.56, 57\%$ Female) is a subset from the American National Election Survey (ANES) from the 2008-2009 Panel Study Off Wave Questionnaires, December 2009 (DeBell, et al, 2010; <https://electionstudies.org/data-center/2008-2009-panel-study/>). All items measure moral preferences and are positively scored of Likert type, ranging from 1 = *Strongly Disagree* to 6 = *Strongly Agree*. Demographic variables include age and gender (male/female). For more details on the `D3mirt` approach, see the vignette included in the package documentation.

# Model Identification

As a first step in the analysis, the three-dimensional GRM must be identified (Reckase, 2009). In the three-dimensional case, this implies locating the x and y-axis.

If the model is not known, this can be done with exploratory factor analysis (EFA), preferably with the help of the EFA option in `mirt::mirt` (Chalmers, 2012) using `ìtemtype = 'graded'`. Note, the EFA model itself is discarded after this step in the procedure.

```{r, message = FALSE, results = 'hide'}
# Load data
data("anes08_09offwaves")
x <- anes08_09offwaves
x <- x[,3:22] # Remove columns for age and gender

# Fit a three-factor EFA model with the mirt package
f <- mirt::mirt(x, 3, itemtype = 'graded')

# Assign data frame with factor loadings with oblimin rotation
g <- summary(f, rotate= 'oblimin')
h <- data.frame(g$rotF)
```

The item lists have one column for the loadings from each item on the factor of interest, and one column with absolute sum scores (denoted `ABS`) for each item calculated from the remaining factor loadings in the model. Each item list is sorted with the lowest absolute sum score highest up.

```{r}
# Call to modid()
modid(h)
```

The top items in each list should be used to identify the model. The first item, the top item in `item.1` located in `list 1`, will be used set the x-axis. In this case, the best item for the x-axis is "W7Q3". The second item, the top item in `Item.2` in `list 2` will be used to identify the y-axis. In this case, the best item for the y-axis is "W7Q20".

# The `D3mirt()` Function

The `D3mirt()` function takes in a data frame with model parameters from a three-dimensional GRM model, specified with orthogonal factors, and set so that all items load on all factors (see examples below). The three-dimensional GRM estimation is preferably done with the `mirt::mirt` (Chalmers, 2012) function. Please note very carefully how the GRM model is specified in the example below.

```{r, message = FALSE, results = 'hide'}
# Load data
data("anes08_09offwaves")
x <- anes08_09offwaves
x <- x[,3:22] # Remove columns for age and gender

# Fit a three-dimensional graded response model with orthogonal factors
# Example below uses Likert items from the built-in data set "anes08_09offwaves"
# Item W7Q3 and item W7Q20 was selected with `modid()`
# The model specification specify all items in the data set (1-20) to load on all three factors (F1-F3)
# The START and FIXED commands are used on the two items to identify the DMIRT model
 spec <- ' F1 = 1-20
           F2 = 1-20
           F3 = 1-20

           START=(W7Q3,a2,0)
           START=(W7Q3,a3,0)

           START=(W7Q20,a3,0)

           FIXED=(W7Q3,a2)
           FIXED=(W7Q3,a3)

           FIXED=(W7Q20,a3) '


mod1 <- mirt::mirt(x, 
                   spec, 
                   itemtype = 'graded', 
                   SE = TRUE, 
                   method = 'QMCEM')

# Assign a data frame with factor loadings (located in the first three columns in mod1),
# and difficulty parameters (columns 4-8 in mod1) with mirt::coef and $'items'[,1:8]))
d <- data.frame(mirt::coef(mod1, 
                           simplify=TRUE)$'items'[,1:8])
```

The `D3mirt()` function call is straightforward. The output, however, is lengthy so it is recommended to use the summary function when inspecting the results.

```{r}
# Call D3mirt() with data frame d
g <- D3mirt(d)
summary(g) # Show summary of results
```

The function then returns an $S3$ object of class `D3mirt` with lists of $a$ and $d$ parameters from the GRM, as well as $MDISC$, $MDIFF$, direction cosines, and degrees for vector angles, construct lists, and vector coordinates assessed by `D3mirt()`.

# The `plotD3mirt` Function

The `plotD3mirt` function is based on the `rgl` package (Adler & Murdoch, 2023) for visualization with OpenGL. Graphing in default mode by calling `plotd3mirt` will return an RGL device that will appear in an external window as a three-dimensional object that can be rotated. In this illustration, however, all RGL devices are shown inline interactive objects. The `view` argument is used in the function call to `plotD3mirt` below but this was only done for optimizing the size of the RGL device (default `view = c(15, 20, 7)`) when creating this R Markdown document.

```{r, testg1, webgl = TRUE, fig.width = 7, fig.height = 7}
# Plot RGL device
plotD3mirt(g, view = c(15, 20, 0.6))
```

The graphical device indicates the presence of foremost two latent dimensions, one aligned with the x-axis and one approaching the y-axis.
The two dimensions, however, have some within-multidimensional tendencies since they are not completely orthogonal.
This tendency, however, is much more prominent for the items W7Q11, W7Q12, W7Q13, and W7Q14 since they are located at around the $45°$ angle or higher between the x and z-axis.

# `D3mirt` Profile Analysis

The `plotD3mirt()` function can also display respondent scores in the three-dimensional model represented as spheres located with the help of factors scores used as coordinates. This allows for a profile analysis in which respondent rows are selected conditioned on some external criteria and then plotted. The resulting output shows where the respondent are located in the model.

To do this, the user must first extract respondent factor scores with `mirt::fscores` (Chalmers, 2012) and then select a subset of respondent rows. The example below select respondent scores by separating on the gender variable

```{r}
# Extract respondent factor scores from mod1 with fscores() function from [mirt::mirt]
f <- mirt::fscores(mod1, 
                   method="EAP", 
                   full.scores = TRUE, 
                   full.scores.SE = F, QMC = T)

# Attach f to gender variable (column 2 from anes08_09offwaves data set; "W3XGENDER")
# Use cbind with fscores() output first
data("anes08_09offwaves")
x <- anes08_09offwaves
z <- data.frame(cbind(f, x[,2]))
```

The `plotD3mirt`function uses `as.factor()` to count the number of factor levels in the data imputed in the `levels` argument. This means that raw data can be used but the number of colors in the color vectors argument (`sphere.col`) may need to be adapted. In the example below, the criteria variable for gender only hold two factor levels. Therefore only two colors in the color vector are needed. The three axes have also been named in the function call.

```{r, testg2, webgl=TRUE, fig.width = 7, fig.height = 7}
# Plot profiles with item vector arrows hidden with hide = TRUE
# Score levels: 1 = Blue ("male") and 2 = Red ("female")
plotD3mirt(g, hide = TRUE, 
           profiles = z, 
           levels = z[,4], 
           sphere.col = c("blue", "red"), 
           x.lab = "Compassion", 
           y.lab="Conformity", 
           z.lab="Fairness", 
           view = c(15, 20, 0.6))
```

In the graphical output, it can be observed a simple profile on gender in which more women tend to have higher levels of trait on compassion.
There seems not to be any observable gender difference related to Conformity or Fairness.

# Exporting The RGL Device

Some options for exporting the RGL device are shown below. Over and above these, it is also possible to export graphical devices in R Markdown documents with `rgl::hookwebgl()` together with graphical options for knitr, as was done in this vignette.

```{r, eval = FALSE}
# Export an open RGL device to the console to be saved as html or image file
plotD3mirt(g, 
           constructs = TRUE)
s <- scene3d()
rgl::rglwidget(s, 
               width = 1040, 
               height = 1040)

# Export a snap shoot of an open RGL device directly to file
plotD3mirt(g, 
           constructs = TRUE)
rgl::rgl.snapshot3d('RGLdevice.png', 
                    fmt = 'png')
```

# Getting help

If you encounter a bug, please file an issue with a minimal reproducible example on GitHub (https://github.com/ForsbergPyschometrics/D3mirt). For questions please contact me on Github or via email (forsbergpsychometrics@gmail.com).
# References

Adler, D., & Murdoch, D. (2023). *Rgl: 3d Visualization Using OpenGL* [Computer software]. <https://dmurdoch.github.io/rgl/index.html>

Chalmers, R., P. (2012). mirt: A Multidimensional Item Response Theory Package for the R Environment. *Journal of Statistical Software, 48*(6), 1-29.

DeBell, M., Krosnick, J. A., & Lupia, A.(2010). *Methodology Report and User's Guide for the 2008--2009 ANES Panel Study*. Palo Alto, CA, and Ann Arbor, MI: Stanford University and the University of Michigan.

Reckase, M. D.(2009).*Multidimensional Item Response Theory*.Springer.

Reckase, M. D.(1985). The Difficulty of Test Items That Measure More Than One Ability. *Applied Psychological Measurement, 9*(4),401-412.<https://doi-org.ezp.sub.su.se/10.1177/014662168500900409>

Reckase, M. D., & McKinley, R. L. (1991). The Discriminating Power of Items That Measure More Than One Dimension. *Applied Psychological Measurement, 15*(4), 361-373. <https://doi-org.ezp.sub.su.se/10.1177/014662169101500407>

Samejima, F. (1969). Estimation of latent ability using a response pattern of graded scores. *Psychometrika 34*, 1-97.<https://doi.org/10.1007/BF03372160>
