---
title: "Introduction to D3mirt Analysis"
author: "Erik Forsberg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to D3mirt Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(D3mirt)
library(mirt)
knitr::knit_hooks$set(webgl = hook_webgl)
```

# `D3MIRT` Modeling

The `D3mirt` analysis is based on descriptive multidimensional item response theory (DMIRT; Reckase, 2009, 1985, Reckase & McKinley, 1991) that can be used on dichotomous and polytomous items.
In DMIRT analysis, also called within multidimensional modeling, it is assumed that items in a multidimensional trait space can measure single or multiple latent abilities (Reckase, 2009, 1985, Reckase & McKinley, 1991).
The DMIRT method is said to be descriptive because the estimates describes item characteristics when more than one latent dimension is used in the analysis.

Under the assumption of within-dimensionality, the two-parameter graded response model (GRM; Samejima, 1969) is used to extract two-parameter multidimensional item characteristics prior the DMIRT analysis.
For `D3mirt`, this implies that a three-dimensional GRM must be fitted so that discrimination and difficulty parameters for a three-dimensional model can be extracted.
The `D3mirt()`function then calculates all necessary estimates.
Of these, the most central estimates are the single multidimensional discrimination ($MDISC$) parameter and the multidimensional difficulty ($MDIFF$) index for each item.

The $MDISC$ $A_i$ for item $i$ represents the highest level of discrimination the item $i$ can achieve located in a multidimensional theta space, with $m$ number of dimensions and $a_{ik}$ item slope parameters (Reckase, 2009).

$$MDISC=A_i=\sqrt{\sum_{k=1}^m a^2_{ik}}$$

Just as in unidimensional modeling, the $A_i$ indicates the direction, as seen from the origin of the model, to the point of maximum slope of the item response surface.
The slope is, similarly to the unidimensional case, assessed as $\frac{A_i}{4}$ (omitted in the equation above).

The items angle orientation is set by taking the direction cosines, in linear algebra terms, of $a_{il}$, i.e., the slope values of item $i$ on coordinate axis $l$ (Reckase, 2009).

$${a_{il}= cos^{-1}\left(\frac{a_{il}}{\sqrt{\sum_{k=1}^m a^2_{ik}}}\right)}$$

The resulting direction vector is a characteristic of the item that describes the angular orientation of an item in a multidimensional theta space.

The multidimensional version of the difficulty parameter, $B_i$, for item $i$ is defined as the negative intercept $d_i$ divided by the $MDISC$ (Reckase, 2009).

$$MDIFF=B_i=\frac{-d_i}{\sqrt{\sum_{k=1}^m a^2_{ik}}}$$

The $MDIFF$ is interpreted similarly as the difficulty parameter in the unidimensional model.
That is, higher values indicate that higher levels of ability for a probability of a correct response of more than .5 are necessary.
Moreover, the $MDIFF$, just as in the unidimensional model, sets the distance along the x-axis from the origin of the model to the point of maximum slope.
However, in DMIRT analysis, the MDIFF becomes a multidimensional location parameter that indicate the distance from the origin to the point of maximum slope following the direction given by the $a_{il}$ equation.
In addition, for Likert items that hold multiple item response functions, the $MDIFF$ can, therefore, be said to show the multidimensional range of difficulty for an item.

Lastly, the $MDISC$ is visualized by scaling the location given the $MDIFF$ and $a_{il}$ so that items with higher discrimination have longer vector arrows (Reckase, 2009).
This also implies that shorter vector arrows indicates lower discrimination that in turn indicate increased amount of model violations of the GRM.

In what follows, the `D3mirt` procedure will be described by function on the built in data set "anes08_09offwaves".
The data ($N = 1046, Mean_age = 51.33, SD = 14.56, 57% Female$) is a subset from the American National Election Survey (ANES) from the ANES 2008-2009 Panel Study Off Wave Questionnaires, December 2009 (DeBell,et al, 2010; <https://electionstudies.org/data-center/2008-2009-panel-study/>).
All items measure moral preferences and are positively scored of Likert type, ranging from $1 = Strongly \, Disagree$ to $6 = Strongly \, Agree$.
Demographic variables include age and gender (male/female).

## Limitations

The `D3mirt` analysis technique is based on the GRM extended to a three-dimensional space.
Consequently, the latter entails that `D3mirt` analysis is limited to items that fit the GRM.
However, the number of dimensions need not be exactly three, but up to three.
This since the third axis, the z-axis, is free while only two items are used for the model identification, i.e., to locate the first (x-axis) and second axis (y-axis).
Thus, if two items can be found that can identify the model, the analysis can be used.

# The `modid()` Function

## Model Identification

As a first step in the `D3mirt` approach, the three-dimensional GRM must be identified (Reckase, 2009).
This implies choosing and fixing two items from the scale or set, and specify the model so that all items load on all factors in the model.
Regarding identification of the three-dimensional GRM, the first item is constrained not to load on the second and third axes (y and z), while the second item is only constrained not to load on the third axis (z).
This will create an orthogonal structure with the first item locked in parallel on the x-axis.
The `modid()` function can help with this by suggesting what items to use for this purpose.

Importantly, if improper items are chosen the model will be hard to interpret meaningfully.
Because of this, the `modid` function was developed to maximize the interpretive meaning analytically by first order factors by sum of squares and the from this select the most optimal items.
This order the model so that the strongest loading items, from the strongest factor, always align with the x-axis.
The model identification process is briefly explained below.

## Step 1: Explore Data Structure

To begin the factor structure must be explored with exploratory factor analysis (EFA).
However, because `D3mirt` analysis is based on item response theory, it is recommended to use multidimensional item response theory EFA methods, such as the EFA option in `mirt::mirt` (Chalmers, 2012) with `ìtemtype = 'graded'`, so that the EFA is performed using the GRM as the item model.
Note, the EFA is only used to find model identification items that meet the necessary DMIRT requirements.
The model itself is discarded after this step in the procedure.

```{r}
# Load data
data("anes08_09offwaves")
x <- anes08_09offwaves
x <- x[,3:22] # Remove columns for age and sex

# Fit a three-factor EFA model with mirt package
f <- mirt::mirt(x, 3, itemtype = 'graded')

# Assign data frame with factor loadings with oblimin rotation
g <- summary(f, rotate= 'oblimin')
h <- data.frame(g$rotF)
```

## Step 2: Item Selection

The `modid()` takes in the factor solution from the EFA, assigned to a data frame $x$, and outputs an S3 object of class `modid` containing list of estimates.
The most important list are the item lists, (denoted $\$items$), with suggestions of what items ($item.1...item.n$) to use for the model identification.
The item lists have one column for the loadings from each item on the factor of interest, and one column with absolute sum scores for each item calculated from the remaining factor loadings in the model.
Each item list is sorted with the lowest absolute sum score highest up.
Accordingly, the top items in each list are the items that best meet the assumption of orthogonality in the analysis.
Therefore, for a three-dimensional model, all else equal, the item highest up in the first list should be used to identify the x-axis, and the item highest up in the second list should be used to identify the y-axis, and so on

```{r}
# Call to modid
modid(h)
```

As can be seen, the first item, used to set the x-axis, is found in the first list, list 1 and "item.1".
In this case the best item for the x-axis is "W7Q3".
The item that identify the y-axis is found in the next list, list 2 and "Item.2".
In this case, the best item for the y-axis is "W7Q20".

Sometimes, however, the model is harder to identify.
If this happens, the first option is to change the rotation method in the EFA, e.g., to change from $oblimin$ to $varimax$.
If this does not solve the issue, the user can try adjusting the `upper`and `lower`arguments in `modid()`.
the latter argument, however, should be the last resort.
This because the upper bound sets the upper limit for inclusion.
Setting this limit to high implies a risk for analyzing data that does not meet the necessary statistical requirements.
The lower limit, however, only increase the size of the item pool used for the item selection.
The user is, therefore, recommended to adjust the lower limit up and down to see if the output differs, and from that make the final decision on what items to use.

```{r}
#' # Call to modid with increased lower and higher bound
modid(h, lower = 1, upper = .12 )
```

In this case the same items are suggested even after adjusting both the `lower`and `upper` limit.

Another option is to override the factor order with the `fac.order` argument.
More specifically, `modid` orders factor by squared factor loadings so that the strongest factor is used first, the second strongest factor is used second, and so on.
Sometimes however, there is only a very small difference between the squared factor loading.
In such situations it can be useful to rearrange the factor order manually to see if the model solution changes (foremost observable when plotting).

```{r}
# Override factor order by reversing columns in the original data frame
modid(h, fac.order = c(3,2,1))
```

In this case, we find the item that was previously suggested for the y-axis, to be suggested for the x-axis.
For the y-axis, however, we now find a new item, "W7Q13", to identify the y-axis.
Since the squared factor loadings for factor 2 and 3 are quite similar, it can be useful to try this suggestion and then compare the results.

## The Model Identification Procedure

Some details on the `modid()` function and its algorithm.
The function uses an iterating procedure that can be user adjusted.
In brief, in the default automatic mode, `modid()` starts by first calculating the ss loadings on all factors $F$ in the data frame $x$ and then rearrange the columns in $x$, in decreasing order following the level of strength of the sum of squares loadings.
Next, the function creates a list containing factor loadings on the first factor, $f_1$, and absolute sum scores of the factor loadings in the remaining factors, i.e., $F-f_1$, row-wise.
The list is then rearrange in decreasing order based on factor loading strength on $f_1$.
Items are selected by scaling $f_1$, and using a standard deviation of $0.5$ (can be adjusted with the `lower` argument.) as the lower bound criteria for inclusion.
That is, starting form the top, rows with raw factor scores and absolut sum scores are extracted until the lower bound is reached.
This allows the function to extract more rows in the case empirical factor loadings are similar in strength.
The result is recorded as a nested list before the function starts over with the next factor, $f_2$, and so on.

For every iteration, the algorithm jumps to the next factor in the EFA model, rearrange rows and extract the strongest loading items.
However the absolute sum score is always assessed on the number of factors less than the total number of factors following the order of iteration, That is, iteration $1$ use factor loadings from all factors $F-f_1$, iteration \$2 $F-(f_{1-2})$, iteration $3$ $F-(f_{1-3})$, and so on, when calculating the absolute sum scores.

## Criteria For Model Identification

Optimized model identification items should preferably (a) have an absolute sum score of $\leq .10$ and (b) with maximized factor loading on the factor of interest.
Of these two criteria, (a) should be given the strongest weight in the selection decision.
If these conditions cannot be met, the user is advised to proceed with caution since the loading scores imply that an adequate orthogonal structure may not be empirically attainable.
If problems in the model identification process occur, please follow the advice given above.

## Limitations

The `modid()` function is not limited to three-dimensional analysis and can be used on any number of factors.
Although based on suggestions on model identification given by Reckase (2009) for this type of analysis, the function offers some expansions that introduce more precision.
The latter foremost consist in incorporating sum of squares in the item selection process (unless the user has not specified otherwise).
Experience tells that this is good practice that often leads to better results compared to other options.
However, it is important to recognize that the model identification procedure only gives suggestions to the model specification, and there could be situations where the researcher should consider other methods.

# The `D3mirt()` Function

The Dmirt function takes in an data frame with model parameters from a three-dimensional GRM model specified with all items loading on all factors (see examples below).
The function then returns an S3 object of class `D3mirt` with lists of $a$ and $d$ parameters from the imputed GRM, $MDISC$, $MDIFF$, direction cosines and degrees for vector angles, construct lists, and vector coordinates.

To initiate the estimation process, the GRM must first be estimated.
This is done with the `mirt::mirt`function.
Please note and copy very carefully how the GRM model is specified in the example below.

```{r}
# Load data
data("anes08_09offwaves")
x <- anes08_09offwaves
x <- x[,3:22] # Remove columns for age and gender

# Fit a three-dimensional graded response model with orthogonal factors
# Example below use Likert items from the built in data set "anes08_09offwaves"
# Item W7Q3 and item W7Q20 was selected with `modid()`
# The model specification specify all items (1-20) to load on all three factors (F1-F3)
# The START and FIXED commands are used to identify the orthogonal structure in the model
 spec <- ' F1 = 1-20
           F2 = 1-20
           F3 = 1-20

           START=(W7Q3,a2,0)
           START=(W7Q3,a3,0)

           START=(W7Q20,a3,0)

           FIXED=(W7Q3,a2)
           FIXED=(W7Q3,a3)

           FIXED=(W7Q20,a3) '


mod1 <- mirt::mirt(x, spec, itemtype = 'graded', SE = TRUE, method = 'QMCEM')

# Assign data frame with factor loadings (located in the first three columns in mod1),
# and difficulty parameters (columns 4-7 in mod1) with mirt::coef and $'items'[,1:7]))
d <- data.frame(mirt::coef(mod1, simplify=TRUE)$'items'[,1:8])
```

The `D3mirt()` function call is straightforward.
The output, however, is lengthy so to use the summary function included in the package to see inspect the results is recommended.

```{r}
# Call `D3mirt()` with data frame d
g <- D3mirt(d)
summary(g) # Show summary of results
```

## Constructs

The user has the option of including constructs in the estimation, by creating one or more nested lists that indicate what items belong to what construct.
From this, the `D3mirt()` function calculates direction cosines for the constructs by adding and normalizing the direction cosines for the items contained in each construct list.
The construct vector arrows can contribute to the analysis by visualizing the average direction for a subset set of items.

```{r}
# Call to `D3mirt()`, including optional nested lists for two constructs
# Item W7Q16 is not included in any construct because of model violations
# The model violations will be possible to see when graphing the model
c <- list(list (1,2,3,4), list(5,7,8,9,10), list(11,12,13,14,15,15,16,17,18,19,20))
g <- D3mirt(d, c)
summary(g)
```

# The `plotD3mirt` Function

The `plotD3mirt` function is based on the `rgl` package for visualization with OpenGL.
The output consists of a three-dimensional interactive RGL device displaying vector arrows estimated with the `D3mirt()` function with the latent dimensions running along the orthogonal axes centered at zero.
If polytomous items are used, such as Likert items, each item will have multiple item response functions that run successively along the same angle in the model.

According to DMIRT (Reckase, 2009, 1985, Reckase & McKinley, 1991), the angle of the vector arrows, seen from the model axes, indicates the direction of maximal slope of discrimination for the particular item response function.
In turn, this also indicates what singular traits, located along the orthogonal axes, an item can be said to describe.
For instance, an item located at $0°$ seen from x-axis, and $90°$ as seen from the $y$ and z-axis, only describes trait $x$.
Such an item is unidimensional because its direction vector of maximal discrimination slope lies parallel and on the x-axis.
In contrast, an item located at $45°$ between all three axes in a three-dimensional model describes all three traits in the model equally well.
Such an item is within-multidimensional because its direction vector lies parallel and on the $45°$ degree line.

## Model Violations

Since DMIRT is based on the multidimensional version of the GRM, all items must adequately meet the necessary statistical assumptions of this type of item model.
Thus, when plotting the `D3mirt` with `plotD3mirt()`model it is possible to visually observe violations of GRM in the graphical device returned.
For instance, shorter vector arrows indicate weaker discrimination and therefore also higher amounts of model violations in comparison.
Moreover, if an item struggles or even fail to describe any of the latent variables in the model, it can often lead to an extreme stretch of the $MDIFF$ range.
This is comparable to trace lines turning horizontal in a unidimensional item response theory model.
Some examples of this will be given below.

Graph in default mode by calling `plotd3mirt` using the returned object from the `D3mirt()` function of class `D3mirt`.
The RGL device will appear in an external window in the noraml case.
In this illustration, however, all RGL devices are shown inline below.

```{r, testg1, webgl=TRUE, echo=FALSE,  fig.align = 'center'}
# Plot RGL device
plotD3mirt(g)
```

The user has the option of plotting the model on one level of difficulty at at time with the `diff.level`argument.
Note, difficulty is the number of response options minus $1$.
In this case, $6$ options were used which means that the model has $5$ levels of difficulty.

```{r, testg2, webgl=TRUE, echo=FALSE, fig.align = 'center'}
# Plot RGL device on one level of difficulty
plotD3mirt(g, diff.level = 5)
```

To plot constructs, change the set `constructs = TRUE`.
The constructs are named with strings as input to the `construct.lab` argument.

```{r, testg3, webgl=TRUE, echo=FALSE,  fig.align = 'center'}
# Plot RGL device with constructs visible and named
plotD3mirt(g, constructs = TRUE, construct.lab = c("Fairness", "Conformity", "Compassion"))
```

A subset of items can be plotted for more thorough investigation.
In the example below constructs are kept which can help with the visual interpretation of the model.

```{r, testg4, webgl=TRUE, echo=FALSE,  fig.align = 'center'}
# A selection of conformity items from the model plotted with constructs
plotD3mirt(g, constructs = TRUE, items = c(5,7,8,9,10), construct.lab = c("Fairness", "Conformity", "Compassion"))
```

## Scaling of Item Vector Arrows

The `D3mirt()` function returns vector coordinates estimated with and without the $MDISC$ as a scalar for the arrow length.
When the MDISC is not used for the arrow length, all vector arrows are scaled to one unit length.
This allows the user to graph the vector arrows with `plotD3mirt()` set to equal length, or more specifically, scaled to one unit length.
This can help reduce clutter in the graphical output when using `plotD3mirt()`.

```{r, testg5, webgl=TRUE, echo=FALSE,  fig.align = 'center'}
# Plot RGL device with scaled items and constructs visible and named
plotD3mirt(g, scale = TRUE, constructs = TRUE, construct.lab = c("Fairness", "Conformity", "Compassion"))
```

## Profile analysis

```{r}
# Plot respondents scores separating on gender variable
# Extract respondent factor scores from mod1 with `fscores()` function from [mirt::mirt]
f <- mirt::fscores(mod1, method="EAP", full.scores = TRUE, full.scores.SE = F, QMC = T)

# Attach f to gender variable (column 2 from anes08_09offwaves data set; "W3XGENDER")
# Use cbind with `fscores()` output first
data("anes08_09offwaves")
x <- anes08_09offwaves
z <- data.frame(cbind(f, x[,2]))
```

```{r, testg6, webgl=TRUE, echo=FALSE, fig.align = 'center'}
# Plot profiles with item vector arrows hidden
# Score levels: 1 = Blue ("male") and 2 = Red ("female")
plotD3mirt(g, hide = TRUE, profiles = z, levels = z[,4], sphere.col = c("blue", "red"), x.lab = "Compassion", y.lab="Conformity", z.lab="Fairness")
```

```{r}
# The use of `rep()`makes it possible to create groups based on factor levels
# This example compares respondents 30 years or younger against 70 years or older
# Column bind `fscores()` with age variable ("W3Xage")
y <- data.frame(cbind(f, x[,1]))

# Subset data frame y conditioned on age <= 30
z1 <- subset(y, y[,4] <= 30)

# Subset data frame y conditioned on age >= 70
z2 <- subset(y, y[,4] >= 70)

# Row bind z1 and z2
z <- rbind(z1,z2)

# Check number of factor leveles with `nlevels()` and `as.factor()`
nlevels(as.factor(z1[,4]))
nlevels(as.factor(z2[,4]))

# Use `rep()`to create a color vector to color groups based on the `nlevels()` output
# z1 has 14 factor levels and z2 has 16 factor levels
# z1 respondents are colored black and z2 are colored grey
colvec <- c(rep("black", 14), rep("grey40", 16))
```

```{r, testg7, webgl=TRUE, echo=FALSE, fig.align = 'center'}
# Call plotD3mirt with profile data on age with item vector arrows hidden
plotD3mirt(g, hide = TRUE, profiles = z, levels = z[,4], sphere.col = colvec, x.lab = "Compassion", y.lab="Conformity", z.lab="Fairness")
```

# Exporting The RGL Device
Options for exporting the RGL deviced are shown below.
Over and above these, it is also possible to export graphical devices in rmarkdown documents with `rgl::hookwebgl()` together with graphical options for knitr, as was done in this document.

```{r}
# Export an open RGL device to console that can be saved as html
plotD3mirt(g, constructs = TRUE)
s <- scene3d()
rgl::rglwidget(s, width = 1040, height = 1040)
```

```{r}
# Export a snap shoot of an open RGL device to file
plotD3mirt(g, constructs = TRUE)
rgl::rgl.snapshot('RGLdevice.png', fmt = 'png')
```

# References
Adler, Daniel, and Duncan Murdoch. 2022. *italics*Rgl: 3d Visualization Using OpenGL*italics*.

DeBell, M., Krosnick, J. A., & Lupia, A.(2010).*italics*Methodology Report and User's Guide for the 2008--2009 ANES Panel Study*italics*.Palo Alto, CA, and Ann Arbor, MI: Stanford University and the University of Michigan.

Chalmers, R., P. (2012).mirt: A Multidimensional Item Response Theory Package for the R Environment.*italics*Journal of Statistical Software, 48*italics*(6), 1-29.

Reckase, M. D.
(2009).
*italics*Multidimensional Item Response Theory*italics*.Springer.

Reckase, M. D.(1985). The Difficulty of Test Items That Measure More Than One Ability.
*italics*Applied Psychological Measurement, 9*italics*(4),401-412.<https://doi-org.ezp.sub.su.se/10.1177/014662168500900409>

Reckase, M. D., & McKinley, R. L.
(1991).
The Discriminating Power of Items That Measure More Than One Dimension.*italics*Applied Psychological Measurement, 15*italics*(4),361-373.
<https://doi-org.ezp.sub.su.se/10.1177/014662169101500407>

Samejima, F. (1969).Estimation of latent ability using a response pattern of graded scores.*italics*Psychometrika 34*italics*, 1-97.<https://doi.org/10.1007/BF03372160>
